---
title: "Supplementary workflow"
author: "Greg Gloor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: margin=2cm
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/frontiers.csl
output:
    pdf_document:
        keep_tex: true
        fig_caption: true
        toc: true
        toc_depth: 2
        includes:
            in_header: /Users/ggloor/Documents/0_git/templates/header.tex
        pandoc_args: [
            "-V", "classoption=twocolumn"
        ]
---

# About this document

This document is an .Rmd document and can be found at:

github.com/ggloor/templates

The document is a template for two column R markdown. It requires an installation of \LaTeX to work properly. This document can contain interspersed markdown and R code that may be compiled into a pdf document and supports the figures and assertions in the main article. R code is not exposed in the pdf document but is referred to by `R code block` in the text so that the interested reader can work through the example code themselves.

## Reproducing the analysis
From an R command prompt you can compile this document into PDF if you have \LaTeX and pandoc installed:

`rmarkdown::render('two_column.Rmd')` or you can open the file in RStudio and compile in that environment.

## R packages required

We will need the following R packages and add-ons (`R_block_1`).

1. knitr (CRAN)

```{r R_block_1, results="hide", echo=F, message=F, error=F, warnings=F}

# in theory - requires knitr and LaTeX installed
# from a bash prompt:
# R -e "rmarkdown::render('Frontiers_Supplement.Rmd')"
# from within an R terminal:
# rmarkdown::render('Frontiers_supplement.Rmd')
# commands to load libraries and datasets
library(knitr)
```

\clearpage

```{r child = 'Chapter2/transforms.Rmd'}
```

\clearpage

# References
